The Decreasing Orbital Period of WASP-4\lowercase{b}

### 2018-12-06 paper/ms.tex
update lists and acknowledgements
better EB apsidal precession reference

The second process that long-term timing studies should reveal is
rotation of the orbital ellipse within the orbital plane (apsidal
precession).  This effect has been studied extensively in eclipsing
binaries \citep[{\it e.g.},][]{russell_notes_1939,
schwarzschild_structure_1958,borkovits_eclipse_2015}, and has been
explored theoretically for transiting planets
\citep{heyl_using_2007,pal_periastron_2008,jordan_observability_2008,ragozzine_probing_2009}.
However, apsidal precession has yet to be directly observed in hot
Jupiters, and the orbits would need non-zero eccentricity for it to be
detectable.  Though the eccentricities of hot Jupiters are expected to
be small, there are many mechanisms that could pump them away from
circularity (see \S~\ref{sec:apsidal_precession}, and
\citealt{bailey_understanding_2019}).  For hot Jupiters, the apsidal
precession is usually dominated by the quadrupolar distortion of the
planet by the star's tidal force \citep{ragozzine_probing_2009}.  If we
measured apsidal precession in a hot Jupiter system, we could thus use
the observed precession rate to compute the planet's Love number, giving
a third parameter along with the mass and radius with which to describe
the planet's interior structure \citep[{\it e.g.},][who performed a
similar procedure for HAT-P-13b]{batygin_determination_2009}.  Measuring
small eccentricities and understanding their origin would also help us
understand the dynamical histories and interior structures of hot
Jupiters.  \citep[{\it
e.g.},][respectively]{dawson_origins_2018,ibgui_tidal_2010}

@@

Another mechanism to excite the hot Jupiter's eccentricity is the
Kozai-Lidov mechanism \citep{lidov_evolution_1962,kozai_secular_1962}.
In this case, the orbital plane of the outer companion, ``c'', would
need to be inclined relative to that of the hot Jupiter, ``b'', by at
least $\sin^{-1} \sqrt{2/5} \approx 39^\circ$.
\citet{bailey_understanding_2019} point out how the joint requirements
of Kozai-Lidov oscillations with a non-detection of an additional RV
signal can be combined to produce lower and upper limits on the
companion's mass.  For the Kozai-Lidov mechanism to operate, from
\citet{bailey_understanding_2019} Equation 20 we need
\begin{equation}
  M_{\rm c} > 7.7\,M_\oplus
  \times \left( \frac{a_{\rm c}}{a_{\rm b}} \right)^{3/2},
  \label{eq:kozai_bound}
\end{equation}
where $M_{\rm b,c}$ and $a_{\rm b,c}$ are the mass and semi-major axis
of WASP-4b and the hypothetical WASP-4c, and we have assumed WASP-4b's
Love number is $k_{2,{\rm b}}\approx 0.6$.  For the RV signal of the
companion to remain undetected, it would need to be in the residual
$({\rm O-C})_{\rm RV} = 15.2\,{\rm m s}^{-1}$ reported by
\citet{triaud_spin-orbit_2010}.  Again following
\citet{bailey_understanding_2019}, this implies
\begin{align}
  M_{\rm c} &<
  ({\rm O-C})_{\rm RV}
  \left( \frac{M_\star a_b}{G} \right)^{1/2}
  \left( \frac{a_c}{a_b} \right)^{1/2}
  f^{-1/2}
  \nonumber
  \\
  M_{\rm c} &< 
  24.2\,M_\oplus
  \times 
  \left( \frac{a_c}{a_b} \right)^{1/2}
  f^{-1/2},
  \label{eq:rv_bound}
\end{align}
for $f(e_{\rm c},\omega_{\rm c}, i_{\rm c}) \propto \sin^2 i_{\rm c}$
a geometric prefactor that depends on the argument of periastron
$\omega_{\rm c}$ and inclination $i_{\rm c}$ of the exterior companion
(\citealt{bailey_understanding_2019} Equation 23).  Since WASP-4b
transits along the line of sight, $f$ can be arbitrarily small, and
both of the preceding limits can be easily satisfied.  This
mechanism could produce the needed eccentricity. However it would be more
satifying if long-term radial velocity monitoring showed an additional
planet in the system.  The chances of one existing with a mass between
1-13\,$M_{\rm Jup}$ and orbital semi-major axis between 1-20\,AU is
about 50\%, and a linear trend is currently present at $\sim 2\sigma$
\citep{knutson_friends_2014}.

@@

\acknowledgements{
L.G.B.\ acknowledges helpful discussions with F. Dai and V. Van Eylen. 
%
We acknowledge the use of TESS Alert data, which is currently in a beta test
phase, from the TESS Science Office. Funding for the TESS mission is provided
by NASA's Science Mission directorate. This research has made use of the NASA
Exoplanet Archive, which is operated by the California Institute of Technology,
under contract with the National Aeronautics and Space Administration under the
Exoplanet Exploration Program.  This work made use of NASA's Astrophysics Data
System Bibliographic Services.
This research has made use of the VizieR catalogue access tool, CDS,
Strasbourg, France. The original description of the VizieR service was
published in A\&AS 143, 23.
This work has made use of data from the European Space Agency (ESA) mission
{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.
%
\newline
%
\facility{
	TESS \citep{ricker_transiting_2015},
	Gaia \citep{gaia_collaboration_gaia_2016,gaia_collaboration_gaia_2018}}
}
%
\software{
  \texttt{astrobase} \citep{bhatti_astrobase_2018},
  \texttt{astropy} \citep{the_astropy_collaboration_astropy_2018},
  \texttt{astroquery} \citep{astroquery_2018},
  \texttt{BATMAN} \citep{kreidberg_batman_2015},
  \texttt{corner} \citep{corner_2016},
  \texttt{emcee} \citep{foreman-mackey_emcee_2013},
  \texttt{IPython} \citep{perez_2007},
  \texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
  \texttt{numpy} \citep{walt_numpy_2011}, 
  \texttt{pandas} \citep{mckinney-proc-scipy-2010},
  \texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{jones_scipy_2001}.
  }

>>>>

The second process that long-term timing studies should reveal is
rotation of the orbital ellipse within the orbital plane (apsidal
precession).  This effect has been studied extensively in eclipsing
binaries \citep[{\it e.g.},][]{
schwarzschild_structure_1958,torres_accurate_2010,borkovits_eclipse_2015}, and has been
explored theoretically for transiting planets
\citep{heyl_using_2007,pal_periastron_2008,jordan_observability_2008,ragozzine_probing_2009}.
However, apsidal precession has yet to be directly observed in hot
Jupiters, and the orbits would need non-zero eccentricity for it to be
detectable.  Though the eccentricities of hot Jupiters are expected to
be small, there are many mechanisms that could pump them away from
circularity (see \S~\ref{sec:apsidal_precession}, and
\citealt{bailey_understanding_2019}).  For hot Jupiters, the apsidal
precession is usually dominated by the quadrupolar distortion of the
planet by the star's tidal force \citep{ragozzine_probing_2009}.  If we
measured apsidal precession in a hot Jupiter system, we could thus use
the observed precession rate to compute the planet's Love number, giving
a third parameter along with the mass and radius with which to describe
the planet's interior structure \citep[{\it e.g.},][who performed a
similar procedure for HAT-P-13b]{batygin_determination_2009}.  Measuring
small eccentricities and understanding their origin would also help us
understand the dynamical histories and interior structures of hot
Jupiters.  \citep[{\it
e.g.},][respectively]{dawson_origins_2018,ibgui_tidal_2010}

@@

Another mechanism to excite the hot Jupiter's eccentricity is the
Kozai-Lidov mechanism \citep{lidov_evolution_1962,kozai_secular_1962}.
In this case, the orbital plane of the outer companion, ``c'', would
need to be inclined relative to that of the hot Jupiter, ``b'', by at
least $\sin^{-1} \sqrt{2/5} \approx 39^\circ$.
\citet{bailey_understanding_2019} point out how the joint requirements
of Kozai-Lidov oscillations with a non-detection of an additional RV
signal can be combined to produce lower and upper limits on the
companion's mass.  For the Kozai-Lidov mechanism to operate at maximum
efficiency, from
\citet{bailey_understanding_2019} Equation 20 we need
\begin{equation}
  M_{\rm c} > 7.7\,M_\oplus
  \times \left( \frac{a_{\rm c}}{a_{\rm b}} \right)^{3/2},
  \label{eq:kozai_bound}
\end{equation}
where $M_{\rm b,c}$ and $a_{\rm b,c}$ are the mass and semi-major axis
of WASP-4b and the hypothetical WASP-4c, and we have assumed WASP-4b's
Love number is $k_{2,{\rm b}}\approx 0.6$.  For the RV signal of the
companion to remain undetected, it would need to be in the residual
$({\rm O-C})_{\rm RV} = 15.2\,{\rm m s}^{-1}$ reported by
\citet{triaud_spin-orbit_2010}.  Again following
\citet{bailey_understanding_2019}, this implies
\begin{align}
  M_{\rm c} &<
  ({\rm O-C})_{\rm RV}
  \left( \frac{M_\star a_b}{G} \right)^{1/2}
  \left( \frac{a_c}{a_b} \right)^{1/2}
  f^{-1/2}
  \nonumber
  \\
  M_{\rm c} &< 
  24.2\,M_\oplus
  \times 
  \left( \frac{a_c}{a_b} \right)^{1/2}
  f^{-1/2},
  \label{eq:rv_bound}
\end{align}
for $f(e_{\rm c},\omega_{\rm c}, i_{\rm c}) \propto \sin^2 i_{\rm c}$
a geometric prefactor that depends on the argument of periastron
$\omega_{\rm c}$ and inclination $i_{\rm c}$ of the exterior companion
(\citealt{bailey_understanding_2019} Equation 23).  Since WASP-4b
transits along the line of sight, $f$ can be arbitrarily small, and
both of the preceding limits can be easily satisfied.  This
mechanism could produce the needed eccentricity. However it would be more
satifying if long-term radial velocity monitoring showed an additional
planet in the system.  The chances of one existing with a mass between
1-13\,$M_{\rm Jup}$ and orbital semi-major axis between 1-20\,AU is
about 50\%, and a linear trend is currently present at $\sim 2\sigma$
\citep{knutson_friends_2014}.

@@

\acknowledgements{
L.G.B.\ is happy to acknowledge help with dynamics from A.  Bailey,
and useful brainstorming with F. Dai and V. Van Eylen. 
%
We acknowledge the use of TESS Alert data, which is currently in a beta test
phase, from the TESS Science Office. Funding for the TESS mission is provided
by NASA's Science Mission directorate. This research has made use of the NASA
Exoplanet Archive, which is operated by the California Institute of Technology,
under contract with the National Aeronautics and Space Administration under the
Exoplanet Exploration Program.  This work made use of NASA's Astrophysics Data
System Bibliographic Services.
This research has made use of the VizieR catalogue access tool, CDS,
Strasbourg, France. The original description of the VizieR service was
published in A\&AS 143, 23.
This work has made use of data from the European Space Agency (ESA) mission
{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.
%
\newline
%
\facility{
	TESS \citep{ricker_transiting_2015},
	Gaia \citep{gaia_collaboration_gaia_2016,gaia_collaboration_gaia_2018}}
}
%
\software{
  \texttt{astrobase} \citep{bhatti_astrobase_2018},
  \texttt{astropy} \citep{the_astropy_collaboration_astropy_2018},
  \texttt{astroquery} \citep{astroquery_2018},
  \texttt{BATMAN} \citep{kreidberg_batman_2015},
  \texttt{corner} \citep{corner_2016},
  \texttt{emcee} \citep{foreman-mackey_emcee_2013},
  \texttt{IPython} \citep{perez_2007},
  \texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
  \texttt{numpy} \citep{walt_numpy_2011}, 
  \texttt{pandas} \citep{mckinney-proc-scipy-2010},
  \texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{jones_scipy_2001}.
  }
